---
title: "Protein Structure Analysis"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## installing bio3d package
```{r}
#install.packages("bio3d", dependencies=TRUE)
library("bio3d")
pdb <- read.pdb("1tag")
# Alternatively, you can read a PDB file directly from your local file system using the file name (or the full path to the file) as an argument to read.pdb():
pdb <- read.pdb("C:/Users/mirza/Desktop/R_Programming/Data/Protein_Structure_Analysis/1tag.pdb")
#pdb <- read.pdb("/path/to/my/data/myfile.pdb")
pdb
class(pdb)
#To examine the contents of the pdb object in more detail we can use the attributes function:
attributes(pdb)
head(pdb$atom)
pdb$atom[1:2, ]
pdb$atom[1:3, c("resno","resid","elety","x","y","z")]
# Note that individual $atom records can also be accessed like this
pdb$atom$elety[1:5]

# Which allows us to do the following (see Figure 1.)
plot.bio3d(pdb$atom$b[pdb$calpha], sse=pdb, type="l",helix.col="blue",sheet.col="red" ,ylab="B-factor")

```
## atom select

```{r}
#The atom.select() function also returns numeric indices that can be used for accessing desired subsets of the pdb data. For example:
ca.inds <- atom.select(pdb, "calpha")
head(ca.inds)
head( pdb$atom[ca.inds$atom, ] )
# And selected xyz coordinates
head( pdb$xyz[, ca.inds$xyz] )
# Select chain A
a.inds <- atom.select(pdb, chain="A")

# Select C-alphas of chain A
ca.inds <- atom.select(pdb, "calpha", chain="A")

# We can combine multiple selection criteria to return their intersection
cab.inds <- atom.select(pdb, elety=c("CA","CB"), chain="A", resno=10:20)

#The atom.select() function also contain options to inverse the final selection. For example, including the argument inverse=TRUE in the statement below will select all atoms except the water atoms:

# Select all atoms except waters
nowat.inds <- atom.select(pdb, "water", inverse=TRUE)
#The argument verbose=TRUE can be used to print more information on how the selection components are combined:
sele <- atom.select(pdb, "protein", elety=c("N", "CA", "C"), resno=50:60, verbose=T)
select_pdb <- pdb$atom[sele$atom, ] 
pdb$atom$resid[sele$atom]
aa321(pdb$atom$resid[sele$atom])
```
##Write a PDB object
Use the command write.pdb() to output a structure object to a PDB formatted file on your local hard drive. Below we use function atom.select() to select only the backbone atoms, and trim.pdb() to create a new PDB object based on our selection of backbone atoms. Finally we use the function write.pdb() to write the newly generated PDB file containing only the backbone atoms:
```{r}
# Output a backbone only PDB file to disc
b.inds <- atom.select(pdb, "backbone")
backpdb <- trim.pdb(pdb, b.inds)
setwd("C:/Users/mirza/Desktop/R_Programming/Data/Protein_Structure_Analysis")
write.pdb(backpdb, file="4q21_back.pdb")
```
## Side-note:
The selection statement can directly be provided into function trim.pdb(). Alternatively, function atom.select() can also return the resulting trimmed pdb object using the optional argument value=TRUE. See examples below:

```{r}
# Selection statements can be passed directly to trim.pdb()
backpdb <- trim.pdb(pdb, "backbone")

# The 'value=TRUE' option of atom.select() will result in a PDB object being returned
backpdb1 <- atom.select(pdb, "backbone", value=TRUE)
```
Function write.pdb() contains further arguments to change the specific data in the PDB structure object. For example, using argument resno the residue numbers in the PDB object will be changed according to the input values, e.g. for the purpose of renumbering a PDB object (see also clean.pdb() and convert.pdb() functions):
```{r}
# Renumber all residues
write.pdb(backpdb, file="4q21_back2.pdb",resno=backpdb$atom$resno+10)
# Assign chain B to all residues
write.pdb(backpdb, chain="B")
```
## Manipulate a PDB object
Basic functions for concatenating, trimming, splitting, converting, rotating, translating and superposing PDB files are available but often you will want to manipulate PDB objects in a custom way.

Below we provide a basic example of such a manipulation process where we read in a multi-chained PDB structure, reassign chain identifiers, and renumber selected residues.
```{r}
pdb <- read.pdb("4lhy")
pdb
# select chains A, E and F
inds <- atom.select(pdb, chain=c("A", "E", "F"))

# trim PDB to selection
pdb2 <- trim.pdb(pdb, inds)

# assign new chain identifiers
pdb2$atom$chain[ pdb2$atom$chain=="E" ] <- "B"
pdb2$atom$chain[ pdb2$atom$chain=="F" ] <- "C"

# re-number chain B and C
pdb2$atom$resno[ pdb2$atom$chain=="B" ] <- pdb2$atom$resno[ pdb2$atom$chain=="B" ] - 156
pdb2$atom$resno[ pdb2$atom$chain=="C" ] <- pdb2$atom$resno[ pdb2$atom$chain=="C" ] - 156

# assign the GDP residue a residue number of 500
pdb2$atom$resno[ pdb2$atom$resid=="GDP" ] <- 500

# use chain D for the GDP residue
pdb2$atom$chain[ pdb2$atom$resid=="GDP" ] <- "D"

# write the new pdb object to file
write.pdb(pdb2, xyz=xyz, file="4LHY_AEF-oriented.pdb")

```

##Concatenate multiple PDBs
Function cat.pdb() can be used to concatenate two or more PDB files. This function contains moreover arguments to re-assign residue numbers and chain identifiers. In the example below we illustrate how to concatenate 4q21 with specific components of 4lhy into a new PDB object:
```{r}
# read two G-protein structures
a <- read.pdb("4q21")
b <- read.pdb("4lhy")

a1 <- trim.pdb(a, chain="A")

b1 <- trim.pdb(b, chain="A")
b2 <- trim.pdb(b, chain="E")
b3 <- trim.pdb(b, chain="F")

# concatenate PDBs
new <- cat.pdb(a1, b1, b2, b3, rechain=TRUE)
unique(new$atom$chain)

# write new PDB object to file
write.pdb(new, file="4Q21-4LHY.pdb")

```

## Coordinate superposition and structural alignment
Structure superposition is often essential for the direct comparison of multiple structures. Bio3D offers versatile functionality for coordinate superposition at various levels. The simplest level is sequence only based superposition:
```{r}
# Align and superpose two or more structures
#download muscle from http://www.drive5.com/muscle/downloads.htm
pdbs <- pdbaln(c("4q21", "521p"), exefile="C:/Users/mirza/Desktop/R_Programming/muscle",fit=TRUE)

a <- read.pdb("4q21")
b <- read.pdb("4lhy")
aln <- struct.aln(a,b,
           exefile="C:/Users/mirza/Desktop/R_Programming/muscle")
# store new coordinates of protein B
b$xyz <- aln$xyz
```
## Working with multiple PDB files
The Bio3D package was designed to specifically facilitate the analysis of multiple structures from both experiment and simulation. The challenge of working with these structures is that they are usually different in their composition (i.e. contain differing number of atoms, sequences, chains, ligands, structures, conformations etc. even for the same protein as we will see below) and it is these differences that are frequently of most interest.

For this reason Bio3D contains extensive utilities to enable the reading, writing, manipulation and analysis of such heterogenous structure sets. This topic is detailed extensively in the separate Principal Component Analysis vignette available from http://thegrantlab.org/bio3d/tutorials.
Before delving into more advanced analysis (detailed in additional vignettes) lets examine how we can read multiple PDB structures from the RCSB PDB for a particular protein and perform some basic analysis:
```{r}
# Download some example PDB files
ids <- c("1TND_B","1AGR_A","1FQJ_A","1TAG_A","1GG2_A","1KJY_A")
raw.files <- get.pdb(ids)
# Extract and align the chains we are interested in
files <- pdbsplit(raw.files, ids)
pdbs <- pdbaln(files,exefile="C:/Users/mirza/Desktop/R_Programming/muscle")
#Below we examine the sequence and structural similarity.
# Calculate sequence identity
pdbs$id <- basename.pdb(pdbs$id)
seqidentity(pdbs)
## Calculate RMSD
rmsd(pdbs, fit=TRUE)

## Quick PCA (see Figure 9)
pc <- pca(pdbfit(pdbs), rm.gaps=TRUE)
plot(pc)
```